---
title: "03 Build Human MG Reference From Prater Green"
author: "jmk"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

## Overview

This notebook builds the human microglia reference object used in downstream
cross-species mapping analyses. It starts from the Prater/Green PU1 microglia
source dataset, standardizes the Seurat object, preserves cluster labels, and
exports the final reference as `human_MG_reference.rds`.

## Citation

> Prater KE, Green KJ, Mamde S, Sun W, Cochoit A, Smith CL, Chiou KL, Heath L,
> Rose SE, Wiley J, Keene CD, Kwon RY, Snyder-Mackler N, Blue EE, Logsdon B,
> Young JE, Shojaie A, Garden GA, Jayadev S. Human microglia show unique
> transcriptional changes in Alzheimer's disease. Nat Aging. 2023
> Jul;3(7):894-907. doi: 10.1038/s43587-023-00424-y.

## Inputs

- `Synaps_datasets/Pu1/Prater_Green_PU1_MGsubset_10clusters_DeID.rdata`

## Outputs

- `Analysis/Human/human_MG_reference.rds`
- `Analysis/Results/UMAP/human_MG_reference_umap_nb03.svg`
  publication-quality UMAP with legend

## Analysis Scope

1. Locate and load the Prater/Green source object.
2. Update to current Seurat object structure.
3. Ensure PCA and UMAP are present (compute if missing).
4. Preserve cluster identities in metadata as `cluster_id`.
5. Save the reference object for notebooks `04` and `05`.

## Handoff Contract

- Consumes the external Prater/Green human microglia source object
- Produces the canonical human microglia reference used by notebooks `04` and `05`
- Writes durable outputs only under `Human/` and `Results/UMAP/` within the analysis root

## Pipeline Flow

1. Prater Green source object\nrdata or legacy rds
2. Load and standardize Seurat object
3. Validate or compute PCA and UMAP
4. Preserve cluster labels as cluster_id
5. Export canonical human_MG_reference.rds
6. Export publication UMAP
7. Notebook 04 major-population mapping
8. Notebook 05 homeostatic subtyping

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

required_packages <- c("Seurat", "SeuratObject", "ggplot2", "viridis")
missing_pkgs <- required_packages[
  !sapply(required_packages, requireNamespace, quietly = TRUE)
]
if (length(missing_pkgs) > 0) {
  stop("Missing packages: ", paste(missing_pkgs, collapse = ", "))
}

library(Seurat)
library(SeuratObject)
library(ggplot2)
library(viridis)

random_seed <- 42
set.seed(random_seed)

analysis_root <- tryCatch(
  {
    f <- knitr::current_input(dir = TRUE)
    if (!is.null(f) && nzchar(f)) dirname(dirname(f)) else stop("not rendering")
  },
  error = function(e) {
    root <- Sys.getenv("ANALYSIS_ROOT", unset = "")
    if (nzchar(root)) root else {
      wd <- getwd()
      if (basename(wd) == "Notebooks") dirname(wd) else wd
    }
  }
)
stopifnot(dir.exists(analysis_root))

human_dir <- file.path(analysis_root, "Human")
results_dir <- file.path(analysis_root, "Results")
umap_dir <- file.path(results_dir, "UMAP")
dir.create(human_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(umap_dir, recursive = TRUE, showWarnings = FALSE)

notebook_tag <- "nb03"

output_rds    <- file.path(human_dir, "human_MG_reference.rds")
output_umap_svg <- file.path(
  umap_dir,
  paste0("human_MG_reference_umap_", notebook_tag, ".svg")
)
```

# Resolve Source File

```{r resolve-source}
source_candidates <- c(
  file.path(
    dirname(analysis_root),
    "Synaps_datasets",
    "Pu1",
    "Prater_Green_PU1_MGsubset_10clusters_DeID.rdata"
  ),
  file.path(
    analysis_root,
    "Synaps_datasets",
    "Pu1",
    "Prater_Green_PU1_MGsubset_10clusters_DeID.rdata"
  ),
  file.path(analysis_root, "Old_Analysis", "Human", "human_MG_seurat.rds"),
  file.path(
    dirname(analysis_root),
    "Old_Analysis",
    "Human",
    "human_MG_seurat.rds"
  )
)

source_path <- source_candidates[file.exists(source_candidates)][1]
if (is.na(source_path)) {
  stop(
    "No source file found. Checked: ",
    paste(source_candidates, collapse = "; ")
  )
}

message("Using source: ", source_path)
```

# Load and Standardize Seurat Object

```{r load-object}
if (grepl("\\.rdata$", source_path, ignore.case = TRUE)) {
  env <- new.env(parent = emptyenv())
  load(source_path, envir = env)

  object_candidates <- Filter(
    function(nm) inherits(get(nm, envir = env), "Seurat"),
    ls(env)
  )

  preferred <- c("ss_data_norm", "human_MG_seurat", "seurat_obj")
  pick <- preferred[preferred %in% object_candidates][1]
  if (is.na(pick)) pick <- object_candidates[1]

  if (is.na(pick)) {
    stop("No Seurat object found in source .rdata")
  }

  human_mg_ref <- get(pick, envir = env)
  message("Loaded Seurat object from .rdata variable: ", pick)
} else {
  human_mg_ref <- readRDS(source_path)
  if (!inherits(human_mg_ref, "Seurat")) {
    stop("Source RDS does not contain a Seurat object")
  }
  message("Loaded Seurat object from .rds")
}

human_mg_ref <- UpdateSeuratObject(human_mg_ref)
message("Cells: ", ncol(human_mg_ref), " | Genes: ", nrow(human_mg_ref))
```

# Ensure Dimensional Reductions

```{r ensure-reductions}
if (!("pca" %in% names(human_mg_ref@reductions))) {
  message("PCA not found; computing PCA")
  human_mg_ref <- RunPCA(human_mg_ref, npcs = 50, verbose = FALSE)
}

if (!("umap" %in% names(human_mg_ref@reductions))) {
  message("UMAP not found; computing UMAP from PCA")
  human_mg_ref <- RunUMAP(
    human_mg_ref,
    reduction = "pca",
    dims = 1:30,
    seed.use = random_seed
  )
}

print(DimPlot(human_mg_ref, reduction = "umap", label = TRUE) +
        ggtitle("Human MG Reference UMAP"))
```

# Preserve Cluster Labels

```{r preserve-cluster-id}
human_mg_ref$cluster_id <- as.character(Idents(human_mg_ref))
if (!"cluster_id" %in% colnames(human_mg_ref@meta.data)) {
  stop("Failed to create metadata column 'cluster_id'")
}

message(
  "cluster_id levels: ",
  paste(sort(unique(human_mg_ref$cluster_id)), collapse = ", ")
)
```

# Generate Publication-Quality UMAP Figure

```{r plot-umap-publication}
# Extract UMAP coordinates and metadata
umap_embedding <- Embeddings(human_mg_ref, reduction = "umap")
umap_df <- data.frame(
  UMAP_1 = umap_embedding[, 1],
  UMAP_2 = umap_embedding[, 2],
  cluster_id = as.factor(human_mg_ref$cluster_id)
)

# Create ggplot2-based UMAP with publication-quality formatting
p_umap <- ggplot(umap_df, aes(x = UMAP_1, y = UMAP_2, color = cluster_id)) +
  geom_point(size = 0.8, alpha = 0.55) +
  scale_color_viridis_d(
    option = "D",
    begin = 0.18,
    end = 0.82,
    guide = guide_legend(
      nrow = 2,
      byrow = TRUE,
      title = "Cluster ID",
      override.aes = list(size = 3, alpha = 1)
    )
  ) +
  labs(
    title = "Human Microglia Reference UMAP",
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    legend.position = "bottom",
    legend.box = "horizontal",
    plot.margin = margin(10, 10, 10, 10, "pt")
  )

print(p_umap)

# Save as SVG for publication
ggsave(
  filename = output_umap_svg,
  plot = p_umap,
  width = 10,
  height = 8,
  units = "in",
  device = "svg"
)

message("Saved publication-quality UMAP: ", output_umap_svg)
```

# Save Reference Outputs

```{r save-outputs}
# Final safety guard for interactive execution order.
# If earlier chunks were skipped, ensure reductions exist before export.
if (!("pca" %in% names(human_mg_ref@reductions))) {
  message("PCA missing at save step; computing PCA now")
  human_mg_ref <- RunPCA(human_mg_ref, npcs = 50, verbose = FALSE)
}
if (!("umap" %in% names(human_mg_ref@reductions))) {
  message("UMAP missing at save step; computing UMAP now")
  human_mg_ref <- RunUMAP(
    human_mg_ref,
    reduction = "pca",
    dims = 1:30,
    seed.use = random_seed
  )
}

saveRDS(human_mg_ref, file = output_rds)
message("Saved: ", output_rds)
```

## Summary

This notebook builds `human_MG_reference.rds` directly from the Prater/Green
human microglia source object and validates that the output contains:

- a Seurat object updated to current format,
- PCA and UMAP reductions,
- `cluster_id` metadata required by notebook `04`.

Next step: run notebook `04_Human_Mouse_Microglia_Overlap.Rmd` to map mouse
microglia onto this reference.
