---
title: "04 Human-Mouse Major Population Mapping"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

## Overview

Stage 1 typing notebook. This maps mouse microglia onto the full human microglia
reference, assigns major population labels, generates UMAP overlays, and exports
handoff artifacts for stage 2 (`05`).

## Inputs

- `Analysis/Human/human_MG_reference.rds`
- `Analysis/adatas/mouse_microglia_humanized_nb02.h5ad`

## Outputs

- `Analysis/Results/UMAP/human_mouse_aligned_reference_vs_mapped_nb04.svg`
- `Analysis/Results/UMAP/human_mouse_combined_umap_nb04.svg`
- `Analysis/Results/Tables/mouse_major_population_counts_nb04.csv`
- `Analysis/adatas/mouse_microglia_major_mapped_nb04.h5ad`

## Handoff Contract

- Consumes the canonical mouse microglia query produced by notebook `02`
- Consumes the human microglia reference produced by notebook `03`
- Produces the major-population typed mouse handoff used directly by notebook `05`
- Writes all durable outputs under `Results/` and `adatas/` inside the analysis root

## Pipeline Flow

1. Notebook 03\nHuman MG reference
2. Notebook 04\nLoad reference and mouse query
3. Notebook 02\nHumanized mouse microglia h5ad
4. Find transfer anchors\nand map mouse onto human reference
5. Assign major population labels
6. Export mapped mouse object\nnb04 h5ad and rds
7. Export count table and UMAP figures
8. Notebook 05 stage-2\nhomeostatic subtyping


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

required_packages <- c(
  "Seurat", "SeuratObject", "ggplot2", "patchwork", "schard", "scales"
)
missing_pkgs <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_pkgs) > 0) {
  stop("Missing packages: ", paste(missing_pkgs, collapse = ", "))
}

library(Seurat)
library(SeuratObject)
library(ggplot2)
library(patchwork)
library(schard)
library(scales)

analysis_root <- tryCatch(
  {
    f <- knitr::current_input(dir = TRUE)
    if (!is.null(f) && nzchar(f)) dirname(dirname(f)) else stop("not rendering")
  },
  error = function(e) {
    wd <- getwd()
    if (basename(wd) == "Notebooks") dirname(wd) else wd
  }
)
stopifnot(dir.exists(analysis_root))

results_umap_dir <- file.path(analysis_root, "Results", "UMAP")
results_tables_dir <- file.path(analysis_root, "Results", "Tables")
adatas_dir <- file.path(analysis_root, "adatas")
dir.create(results_umap_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(results_tables_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(adatas_dir, recursive = TRUE, showWarnings = FALSE)

notebook_tag <- "nb04"
aligned_plot_path <- file.path(
  results_umap_dir,
  paste0("human_mouse_aligned_reference_vs_mapped_", notebook_tag, ".svg")
)
combined_plot_path <- file.path(
  results_umap_dir,
  paste0("human_mouse_combined_umap_", notebook_tag, ".svg")
)
major_counts_path <- file.path(
  results_tables_dir,
  paste0("mouse_major_population_counts_", notebook_tag, ".csv")
)
major_mapped_h5ad_path <- file.path(
  adatas_dir,
  paste0("mouse_microglia_major_mapped_", notebook_tag, ".h5ad")
)
major_mapped_rds_path <- file.path(
  adatas_dir,
  paste0("mouse_microglia_major_mapped_", notebook_tag, ".rds")
)
```

## Load Data

```{r load-data}
human_ref_path <- file.path(
  analysis_root,
  "Human",
  "human_MG_reference.rds"
)
mouse_query_path <- file.path(
  analysis_root,
  "adatas",
  "mouse_microglia_humanized_nb02.h5ad"
)

if (!file.exists(human_ref_path)) {
  stop("Missing human reference: ", human_ref_path)
}
if (!file.exists(mouse_query_path)) {
  stop("Missing mouse query from nb02: ", mouse_query_path)
}

human_ref <- readRDS(human_ref_path)
mouse_query <- schard::h5ad2seurat(mouse_query_path)

if (!("cluster_id" %in% colnames(human_ref@meta.data))) {
  stop("'cluster_id' is required in human reference metadata")
}
if (!("integrated" %in% names(human_ref@assays))) {
  stop("Assay 'integrated' not found in human reference")
}
```

## Transfer Anchors and Mapping

```{r map-query}
human_scale_data <- tryCatch(
  SeuratObject::LayerData(
    object = human_ref[["integrated"]],
    layer = "scale.data"
  ),
  error = function(e) {
    GetAssayData(object = human_ref[["integrated"]], slot = "scale.data")
  }
)

shared_genes <- intersect(rownames(human_scale_data), rownames(mouse_query))
if (length(shared_genes) < 100) {
  stop("Too few shared genes for mapping: ", length(shared_genes))
}

human_ref_sub <- subset(human_ref, features = shared_genes)
mouse_query_sub <- subset(mouse_query, features = shared_genes)

DefaultAssay(human_ref_sub) <- "integrated"
DefaultAssay(mouse_query_sub) <- "RNA"

if (!("pca" %in% names(human_ref_sub@reductions))) {
  human_ref_sub <- RunPCA(human_ref_sub, npcs = 30, verbose = FALSE)
}

has_umap_model <- function(seurat_obj, reduction_name = "umap") {
  reduction_name %in% names(seurat_obj@reductions) &&
    "model" %in% names(seurat_obj@reductions[[reduction_name]]@misc)
}

if (!has_umap_model(human_ref_sub, "umap")) {
  human_ref_sub <- RunUMAP(
    human_ref_sub,
    reduction = "pca",
    dims = 1:30,
    return.model = TRUE
  )
}

anchors <- FindTransferAnchors(
  reference = human_ref_sub,
  query = mouse_query_sub,
  reference.assay = "integrated",
  query.assay = "RNA",
  reduction = "pcaproject",
  reference.reduction = "pca",
  dims = 1:30
)

mouse_query_mapped <- MapQuery(
  anchorset = anchors,
  query = mouse_query_sub,
  reference = human_ref_sub,
  refdata = list(cluster_id = "cluster_id"),
  reference.reduction = "pca",
  reduction.model = "umap"
)

mouse_query_mapped$major_pop_label <- as.character(
  mouse_query_mapped$predicted.cluster_id
)
```

## Export Major Typing Handoff

```{r export-major-typing}
major_counts <- as.data.frame(
  table(mouse_query_mapped$major_pop_label),
  stringsAsFactors = FALSE
)
colnames(major_counts) <- c("major_pop_label", "n_cells")
major_counts <- major_counts[order(-major_counts$n_cells), ]
write.csv(major_counts, file = major_counts_path, row.names = FALSE)

saveRDS(mouse_query_mapped, major_mapped_rds_path)
message("Saved: ", major_mapped_rds_path)
message("Saved: ", major_counts_path)
```

## UMAP Outputs

```{r umap-outputs}
umap_coords <- Embeddings(human_ref_sub, reduction = "umap")
x_limits <- range(umap_coords[, 1])
y_limits <- range(umap_coords[, 2])

p_human <- DimPlot(
  human_ref_sub,
  reduction = "umap",
  group.by = "cluster_id",
  label = TRUE,
  pt.size = 0.45
) +
  ggtitle("Human Major Microglia Reference") +
  coord_fixed() + xlim(x_limits) + ylim(y_limits)

p_mouse <- DimPlot(
  mouse_query_mapped,
  reduction = "ref.umap",
  group.by = "major_pop_label",
  label = TRUE,
  pt.size = 0.45
) +
  ggtitle("Mapped Mouse Major Populations") +
  coord_fixed() + xlim(x_limits) + ylim(y_limits)

aligned_plot <- p_human + p_mouse
print(aligned_plot)
ggsave(
  filename = aligned_plot_path,
  plot = aligned_plot,
  width = 12,
  height = 6,
  units = "in"
)

human_umap <- Embeddings(human_ref_sub, "umap")
mouse_umap <- Embeddings(mouse_query_mapped, "ref.umap")
combined_df <- data.frame(
  UMAP_1 = c(human_umap[, 1], mouse_umap[, 1]),
  UMAP_2 = c(human_umap[, 2], mouse_umap[, 2]),
  species = c(rep("Human", nrow(human_umap)), rep("Mouse", nrow(mouse_umap))),
  major_pop_label = c(
    as.character(human_ref_sub$cluster_id),
    rep(NA_character_, nrow(mouse_umap))
  )
)

p_combined <- ggplot(combined_df, aes(x = UMAP_1, y = UMAP_2)) +
  geom_point(
    data = subset(combined_df, species == "Human"),
    aes(color = major_pop_label),
    size = 0.4,
    alpha = 0.6
  ) +
  geom_point(
    data = subset(combined_df, species == "Mouse"),
    color = "red",
    size = 2,
    alpha = 0.8
  ) +
  scale_color_hue(h = c(0, 360), c = 60, l = 55, labels = label_wrap(15)) +
  labs(
    title = "Major Population Mapping: Human Reference and Projected Mouse",
    x = "UMAP 1",
    y = "UMAP 2",
    color = "Human major population"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.justification = "left",
    legend.box.just = "left",
    axis.line = element_line(color = "black", linewidth = 0.5)
  ) +
  coord_equal()

print(p_combined)
ggsave(
  filename = combined_plot_path,
  plot = p_combined,
  width = 11,
  height = 8,
  units = "in"
)
```

## Next Step

Run `05_Homeostati_Human_Mouse_Overlap.Rmd` to subtype cells assigned to major
population `1` into homeostatic subpopulations.
