---
title: "05 Homeostatic Subtyping From Major Population 1"
output:
  html_notebook:
    toc: true
    toc_float: true
    number_sections: true
---

## Overview

Production handoff notebook for the second typing event:

1. load nb04 major-typed mouse mapping output,
2. subset mouse cells assigned to major population `1`,
3. map those cells to the human homeostatic reference,
4. export a final combined human+mouse AnnData containing both typing levels.

## Inputs

- `Analysis/adatas/mouse_microglia_major_mapped_nb04.h5ad`
- `Analysis/Human/human_homeostat_ref.rds` (fallback: `human_MG_reference.rds`)

## Outputs

- `Results/UMAP/homeostatic_subtyping_aligned_nb05.svg`
  aligned UMAP for homeostatic subtyping
- `Results/Tables/mouse_homeostatic_subtype_counts_nb05.csv`
  mouse subtype counts within major population 1
- `adatas/human_mouse_homeostatic_typed_nb05.h5ad`
  final handoff with `major_pop_label` and `homeostatic_subtype_label`

## Handoff Contract

- Consumes the major-population mapped mouse object from notebook `04`
- Consumes the human homeostatic reference under `Human/`
- Produces the final cross-species homeostatic typing handoff used directly by notebook `06`
- Writes durable outputs only under `Results/` and `adatas/` within the analysis root

## Pipeline Flow

1. Notebook 04\nMajor-population mapped mouse microglia
2. Load nb04 output
3. Human homeostatic reference\nHuman/human_homeostat_ref.rds\nor human_MG_reference.rds] --> D[
4. Load reference
5. Subset mouse cells\nmajor_pop_label == 1
6. Prepare homeostatic reference
7. Transfer anchors and MapQuery
8. Assign homeostatic subtype labels
9. Export aligned UMAP and subtype counts
10. Export final combined human+mouse h5ad/rds
11. Notebook 06\nCross-species homeostatic analysis

```{r setup, include=FALSE}
# Fail fast so pipeline status reflects real execution outcomes.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, error = FALSE)

required_packages <- c(
  "Seurat", "SeuratObject", "ggplot2", "patchwork", "schard", "Matrix",
  "SingleCellExperiment", "zellkonverter"
)
missing_pkgs <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_pkgs) > 0) {
  stop("Missing packages: ", paste(missing_pkgs, collapse = ", "))
}

library(Seurat)
library(SeuratObject)
library(ggplot2)
library(patchwork)
library(schard)
library(SingleCellExperiment)
library(zellkonverter)

analysis_root <- tryCatch(
  {
    f <- knitr::current_input(dir = TRUE)
    if (!is.null(f) && nzchar(f)) dirname(dirname(f)) else stop("not rendering")
  },
  error = function(e) {
    wd <- getwd()
    if (basename(wd) == "Notebooks") dirname(wd) else wd
  }
)
stopifnot(dir.exists(analysis_root))

results_umap_dir <- file.path(analysis_root, "Results", "UMAP")
results_tables_dir <- file.path(analysis_root, "Results", "Tables")
adatas_dir <- file.path(analysis_root, "adatas")
dir.create(results_umap_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(results_tables_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(adatas_dir, recursive = TRUE, showWarnings = FALSE)

notebook_tag <- "nb05"
input_mouse_major_path <- file.path(
  adatas_dir,
  "mouse_microglia_major_mapped_nb04.h5ad"
)
input_mouse_major_rds_path <- file.path(
  adatas_dir,
  "mouse_microglia_major_mapped_nb04.rds"
)
output_umap_path <- file.path(
  results_umap_dir,
  paste0("homeostatic_subtyping_aligned_", notebook_tag, ".svg")
)
output_counts_path <- file.path(
  results_tables_dir,
  paste0("mouse_homeostatic_subtype_counts_", notebook_tag, ".csv")
)
output_h5ad_path <- file.path(
  adatas_dir,
  paste0("human_mouse_homeostatic_typed_", notebook_tag, ".h5ad")
)
output_rds_path <- file.path(
  adatas_dir,
  paste0("human_mouse_homeostatic_typed_", notebook_tag, ".rds")
)
```

## Load Inputs

```{r load-data}
if (file.exists(input_mouse_major_rds_path)) {
  mouse_major <- readRDS(input_mouse_major_rds_path)
  message("Loaded mouse_major from RDS: ", input_mouse_major_rds_path)
} else if (file.exists(input_mouse_major_path)) {
  mouse_major <- schard::h5ad2seurat(input_mouse_major_path)
  message("Loaded mouse_major from h5ad: ", input_mouse_major_path)
} else {
  stop(
    "Missing required nb04 output. Checked:\n",
    "  RDS: ", input_mouse_major_rds_path, "\n",
    "  h5ad: ", input_mouse_major_path
  )
}

homeostat_ref_candidates <- c(
  file.path(analysis_root, "Human", "human_homeostat_ref.rds"),
  file.path(analysis_root, "Human", "human_MG_reference.rds")
)
existing_homeostat_refs <- homeostat_ref_candidates[file.exists(homeostat_ref_candidates)]
if (length(existing_homeostat_refs) == 0) {
  stop(
    "Missing homeostatic reference. Checked: ",
    paste(homeostat_ref_candidates, collapse = "; ")
  )
}

homeostat_ref <- NULL
homeostat_ref_path <- NA_character_
read_errors <- character(0)
for (candidate in existing_homeostat_refs) {
  ref_try <- tryCatch(readRDS(candidate), error = function(e) e)
  if (!inherits(ref_try, "error")) {
    homeostat_ref <- ref_try
    homeostat_ref_path <- candidate
    break
  }
  read_errors <- c(read_errors, paste0(candidate, " -> ", conditionMessage(ref_try)))
}

if (is.null(homeostat_ref)) {
  stop(
    "Could not read any homeostatic reference RDS. Errors: ",
    paste(read_errors, collapse = " | ")
  )
}

message("Loaded homeostatic reference: ", homeostat_ref_path)

cluster_col <- if ("subcluster_id" %in% colnames(homeostat_ref@meta.data)) {
  "subcluster_id"
} else if ("cluster_id" %in% colnames(homeostat_ref@meta.data)) {
  "cluster_id"
} else {
  stop(
    "Neither 'subcluster_id' nor 'cluster_id' found in homeostatic reference"
  )
}
homeostat_ref$homeostatic_subtype_label <- as.character(
  homeostat_ref[[cluster_col, drop = TRUE]]
)

if (!("major_pop_label" %in% colnames(mouse_major@meta.data))) {
  if ("predicted.cluster_id" %in% colnames(mouse_major@meta.data)) {
    mouse_major$major_pop_label <- as.character(
      mouse_major$predicted.cluster_id
    )
  } else {
    stop("mouse_major is missing major population labels from nb04")
  }
}

message("Loaded mouse cells from nb04: ", ncol(mouse_major))
```

## Subset Major Population 1 and Remap

```{r remap-homeostatic-subtypes}
mouse_major1 <- subset(mouse_major, subset = major_pop_label == "1")
if (ncol(mouse_major1) == 0) {
  stop("No mouse cells with major_pop_label == '1' were found")
}

if (!("integrated" %in% names(homeostat_ref@assays))) {
  stop("Assay 'integrated' not found in homeostatic reference")
}

homeo_genes <- tryCatch(
  rownames(
    SeuratObject::LayerData(
      homeostat_ref[["integrated"]],
      layer = "scale.data"
    )
  ),
  error = function(e) {
    rownames(
      GetAssayData(homeostat_ref, assay = "integrated", slot = "scale.data")
    )
  }
)
shared_genes <- intersect(homeo_genes, rownames(mouse_major1))
if (length(shared_genes) < 100) {
  stop("Too few shared genes for subtype mapping: ", length(shared_genes))
}

homeo_ref_sub <- homeostat_ref[shared_genes, ]
mouse_major1_sub <- mouse_major1[shared_genes, ]

DefaultAssay(homeo_ref_sub) <- "integrated"
DefaultAssay(mouse_major1_sub) <- "RNA"

if (!("pca" %in% names(homeo_ref_sub@reductions))) {
  homeo_ref_sub <- RunPCA(homeo_ref_sub, npcs = 30, verbose = FALSE)
}

has_umap_model <- function(seurat_obj, reduction_name = "umap") {
  reduction_name %in% names(seurat_obj@reductions) &&
    "model" %in% names(seurat_obj@reductions[[reduction_name]]@misc)
}

if (!has_umap_model(homeo_ref_sub, "umap")) {
  homeo_ref_sub <- RunUMAP(
    homeo_ref_sub,
    reduction = "pca",
    dims = 1:30,
    return.model = TRUE
  )
}

anchors <- FindTransferAnchors(
  reference = homeo_ref_sub,
  query = mouse_major1_sub,
  reference.assay = "integrated",
  query.assay = "RNA",
  reduction = "pcaproject",
  reference.reduction = "pca",
  dims = 1:30
)

mouse_major1_mapped <- MapQuery(
  anchorset = anchors,
  query = mouse_major1_sub,
  reference = homeo_ref_sub,
  refdata = list(homeostatic_subtype_label = "homeostatic_subtype_label"),
  reference.reduction = "pca",
  reduction.model = "umap"
)

mouse_major1_mapped$homeostatic_subtype_label <- as.character(
  mouse_major1_mapped$predicted.homeostatic_subtype_label
)
```

## QC UMAP and Counts

```{r qc-outputs}
umap_coords <- Embeddings(homeo_ref_sub, reduction = "umap")
x_limits <- range(umap_coords[, 1])
y_limits <- range(umap_coords[, 2])

all_subtypes <- sort(unique(c(
  as.character(homeo_ref_sub$homeostatic_subtype_label),
  as.character(mouse_major1_mapped$homeostatic_subtype_label)
)))
all_subtypes <- all_subtypes[!is.na(all_subtypes)]
subtype_colors <- setNames(
  grDevices::hcl.colors(length(all_subtypes), palette = "Dark 3"),
  all_subtypes
)

p_human <- DimPlot(
  homeo_ref_sub,
  reduction = "umap",
  group.by = "homeostatic_subtype_label",
  cols = subtype_colors,
  label = TRUE
) +
  ggtitle("Human Homeostatic Subtypes") +
  coord_fixed() + xlim(x_limits) + ylim(y_limits)

p_mouse <- DimPlot(
  mouse_major1_mapped,
  reduction = "ref.umap",
  group.by = "homeostatic_subtype_label",
  cols = subtype_colors,
  label = TRUE
) +
  ggtitle("Mouse Major Pop 1 Mapped to Homeostatic Subtypes") +
  coord_fixed() + xlim(x_limits) + ylim(y_limits)

aligned_plot <- p_human + p_mouse
print(aligned_plot)

svg(output_umap_path, width = 12, height = 6)
print(aligned_plot)
dev.off()
message("Saved: ", output_umap_path)

subtype_counts <- as.data.frame(
  table(mouse_major1_mapped$homeostatic_subtype_label),
  stringsAsFactors = FALSE
)
colnames(subtype_counts) <- c("homeostatic_subtype_label", "n_cells")
subtype_counts <- subtype_counts[order(-subtype_counts$n_cells), ]
write.csv(subtype_counts, file = output_counts_path, row.names = FALSE)
message("Saved: ", output_counts_path)
```

## Export Final Combined Handoff

```{r export-final-handoff}

# --- Harmonize subtypes: only keep human cells with subtypes present in mouse ---
mouse_subtypes <- unique(na.omit(mouse_major1_mapped$homeostatic_subtype_label))
homeo_ref_sub <- subset(
  homeo_ref_sub,
  cells = colnames(homeo_ref_sub)[homeo_ref_sub$homeostatic_subtype_label %in% mouse_subtypes]
)
# Relabel to ensure matching
homeo_ref_sub$homeostatic_subtype_label <- factor(
  as.character(homeo_ref_sub$homeostatic_subtype_label),
  levels = mouse_subtypes
)
mouse_major1_mapped$homeostatic_subtype_label <- factor(
  as.character(mouse_major1_mapped$homeostatic_subtype_label),
  levels = mouse_subtypes
)

# Keep both label levels for Python analysis.
homeo_ref_sub$species <- "Human"
homeo_ref_sub$major_pop_label <- NA_character_
mouse_major1_mapped$species <- "Mouse"
if (!("major_pop_label" %in% colnames(mouse_major1_mapped@meta.data))) {
  mouse_major1_mapped$major_pop_label <- "1"
}

if (!("RNA" %in% names(homeo_ref_sub@assays))) {
  stop("Homeostatic reference missing RNA assay")
}
DefaultAssay(homeo_ref_sub) <- "RNA"
DefaultAssay(mouse_major1_mapped) <- "RNA"

# Preserve counts from RNA assay slots before merge for downstream analysis (e.g. SCANVI)
h_counts <- NULL
m_counts <- NULL
tryCatch(
  {
    h_counts <- GetAssayData(homeo_ref_sub, assay = "RNA", slot = "counts")
    message("  Extracting human RNA counts for export: ",
            nrow(h_counts), " x ", ncol(h_counts))
  },
  error = function(e) {
    message("  Human counts unavailable: ", conditionMessage(e))
  }
)
tryCatch(
  {
    m_counts <- GetAssayData(mouse_major1_mapped, assay = "RNA", slot = "counts")
    message("  Extracting mouse RNA counts for export: ",
            nrow(m_counts), " x ", ncol(m_counts))
  },
  error = function(e) {
    message("  Mouse counts unavailable: ", conditionMessage(e))
  }
)

human_umap <- Embeddings(homeo_ref_sub, reduction = "umap")
mouse_umap <- Embeddings(mouse_major1_mapped, reduction = "ref.umap")
rownames(human_umap) <- paste0("Human_", rownames(human_umap))
rownames(mouse_umap) <- paste0("Mouse_", rownames(mouse_umap))
combined_umap <- rbind(human_umap, mouse_umap)

combined_final <- merge(
  x = homeo_ref_sub,
  y = mouse_major1_mapped,
  add.cell.ids = c("Human", "Mouse"),
  project = "human_mouse_homeostatic_typed"
)

combined_final[["mapped_umap"]] <- CreateDimReducObject(
  embeddings = combined_umap[colnames(combined_final), , drop = FALSE],
  key = "UMAP_",
  assay = DefaultAssay(combined_final)
)

# Merge counts from both sources if available
if (!is.null(h_counts) && !is.null(m_counts)) {
  tryCatch(
    {
      shared_genes_counts <- intersect(rownames(h_counts), rownames(m_counts))
      if (length(shared_genes_counts) > 100) {
        h_c_subset <- h_counts[shared_genes_counts, ]
        m_c_subset <- m_counts[shared_genes_counts, ]
        combined_counts <- cbind(h_c_subset, m_c_subset)
        colnames(combined_counts) <- colnames(combined_final)
        combined_final[["counts"]] <- CreateAssayObject(counts = combined_counts)
        message("  Created combined counts assay: ", 
                length(shared_genes_counts), " shared genes")
      }
    },
    error = function(e) {
      message("  Note: Could not merge counts - ", conditionMessage(e))
    }
  )
}

tryCatch(
  {
    DefaultAssay(combined_final) <- "RNA"
    sce <- as.SingleCellExperiment(combined_final, assay = "RNA")
    
    # If combined_final has a counts assay, add it explicitly to SCE
    if ("counts" %in% names(combined_final@assays)) {
      counts_data <- GetAssayData(combined_final, assay = "counts", slot = "counts")
      SummarizedExperiment::assay(sce, "counts", withDimnames = FALSE) <-
        counts_data[, colnames(sce)]
      message("  Included counts layer in SCE export")
    }
    
    col_data <- as.data.frame(SummarizedExperiment::colData(sce))
    col_data[] <- lapply(
      col_data,
      function(col) {
        if (is.factor(col)) as.character(col) else col
      }
    )
    SummarizedExperiment::colData(sce) <- S4Vectors::DataFrame(col_data)
    row_data <- as.data.frame(SummarizedExperiment::rowData(sce))
    row_data$gene_symbol <- rownames(sce)
    SummarizedExperiment::rowData(sce) <- S4Vectors::DataFrame(row_data)
    SingleCellExperiment::reducedDim(
      sce, "X_umap"
    ) <- combined_umap[colnames(combined_final), , drop = FALSE]
    x_assay <- if ("logcounts" %in% SummarizedExperiment::assayNames(sce)) {
      "logcounts"
    } else if ("counts" %in% SummarizedExperiment::assayNames(sce)) {
      "counts"
    } else {
      SummarizedExperiment::assayNames(sce)[1]
    }
    zellkonverter::writeH5AD(
      sce,
      file = output_h5ad_path,
      X_name = x_assay
    )
    message("Saved h5ad with counts layer: ", output_h5ad_path)
  },
  error = function(e) {
    message("h5ad export failed: ", conditionMessage(e))
    saveRDS(combined_final, output_rds_path)
    message("Saved RDS fallback: ", output_rds_path)
  }
)
```
